### 2025-09-30 causal-inference.qmd
20250930 - butterfly bias
20250930 - variables to control/not to control for

When dealing with a [collider](#sec-colliderBias) that is also a [confound](#sec-causalDiagramConfounding), controlling for either set, `B` and `M` or `A` and `M`, will provide an unbiased estimate of the association between `X` and `Y`.
In this case, controlling for either a) coaching quality and mental health or b) diet and mental health—but not both sets—will yield an unbiased estimate of the association between off-field behavior and fantasy points.

@@

@@

There are three criteria for establishing causality: 1) the cause precedes the effect.
2) The cause is related to the effect.
3) There are no other alternative explanations for the effect apart from the cause.
In general, it is important to be aware of the counterfactual and to consider what would have happened if the supposed cause had not occurred.
Various experimental and quasi-experimental designs and approaches can be leveraged to more closely approximate causal inferences.
[Longitudinal designs](#sec-causalInferenceLongitudinal), [within-subject analyses](#sec-causalInferenceWithinSubject), [inclusion of control variables](#sec-causalInferenceControlVariables), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed) are all quasi-experimental designs that afford the researcher greater control over some possible third variable [confounds](#sec-causalDiagramConfounding).
Causal diagrams can be a useful tool for identifying the proper variables to control for (and those not to control for).
In general, the only variables you should control for are [confound(s)](#sec-causalDiagramConfounding).
When [confounding](#sec-causalDiagramConfounding) exists, it is important to control for the [confound(s)](#sec-causalDiagramConfounding).
It is important not to control for mediators when interested in the total effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).
In addition, it is important not to control for [descendants](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy).
When there is a [collision](#sec-colliderBias), it is important not to control for the [collider](#sec-colliderBias) (unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding)).

>>>>

When dealing with a [collider](#sec-colliderBias) that is also a [confound](#sec-causalDiagramConfounding), controlling for either set, `B` and `M` or `A` and `M`, will provide an unbiased estimate of the association between `X` and `Y`.
In this case, controlling for either a) coaching quality and mental health or b) diet and mental health—but not both sets—will yield an unbiased estimate of the association between off-field behavior and fantasy points.
If you control for the both sets (`M`, `A`, and `B`), you can induce an artificial association between `X` and `Y`, which is called butterfly (or bow-tie) bias.

@@

## Summary of Variables to Control For (and Not to Control For) {#sec-causalInferenceControlVariablesSummary}

In general, you should control for all [confounds](#sec-causalDiagramConfounding).
In addition, it is okay to control for [ancestors](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy) that are not [confounds](#sec-causalDiagramConfounding) or mediators.
That is, it is okay to control for variables that influence `Y` that do not influence `X` and that are not influenced by `X`.
When including these variable as control variables in a model, they are called precision variables.
You do not need to include precision variables in the model because the estimate of the association is already unbiased if you have controlled for all [confounds](#sec-causalDiagramConfounding).
However, including precision variables in the model reduces residual variance in the [outcome variable](#sec-correlationalStudy) and can yield more precise estimates (i.e., smaller standard errors) of the association between the [predictor variable](#sec-correlationalStudy) and [outcome variable](#sec-correlationalStudy).

In addition, there are some variables that are important not to control for.
It is important not to control for mediators of two variables for which you want to determine the estimate of the causal effect—unless you are interested in the *direct* causal effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy) above and beyond the mediator.
In addition, it is is important not to control for a) ancestors of the [predictor variable](#sec-correlationalStudy) that are not [confounds](#sec-causalDiagramConfounding), b) [descendants](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy), and c) [colliders](#sec-colliderBias) (unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding)).

@@

There are three criteria for establishing causality: 1) the cause precedes the effect.
2) The cause is related to the effect.
3) There are no other alternative explanations for the effect apart from the cause.
In general, it is important to be aware of the counterfactual and to consider what would have happened if the supposed cause had not occurred.
Various experimental and quasi-experimental designs and approaches can be leveraged to more closely approximate causal inferences.
[Longitudinal designs](#sec-causalInferenceLongitudinal), [within-subject analyses](#sec-causalInferenceWithinSubject), [inclusion of control variables](#sec-causalInferenceControlVariables), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed) are all quasi-experimental designs that afford the researcher greater control over some possible third variable [confounds](#sec-causalDiagramConfounding).
Causal diagrams can be a useful tool for identifying the proper variables to control for (and those not to control for).
In general, the only variables you should control for are [confounds](#sec-causalDiagramConfounding)—however, it is okay to control for [ancestors](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy) that are not [confounds](#sec-causalDiagramConfounding) or mediators.
When [confounding](#sec-causalDiagramConfounding) exists, it is important to control for the [confound(s)](#sec-causalDiagramConfounding).
It is important not to control for mediators when interested in the total effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).
In addition, it is important not to control for [ancestors](#sec-ancestorsDescendants) of the [predictor variable](#sec-correlationalStudy) or [descendants](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy).
When there is a [collision](#sec-colliderBias), it is important not to control for the [collider](#sec-colliderBias) (unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding)).
